Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-04 summary.tex
appendix.tex

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach to model thermal relaxation was also
employed in the global numerical simulations of \cite{nelson13}, and
allows the thermodynamic response of the disk to be controlled.
However, it is possible to model $\beta$ self-consistently by
relating it to thermal timescales expected in realistic disks.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}, who showed the linear VSI studied by
\cite{nelson13} in a locally isothermal disk can be captured by the
VGSB. We find the VGSB framework is appropriate for $\beta\lesssim1$. 
This is convenient because rapid thermal relaxation ($\beta\ll1$) is
generally required for the VSI.  However, the VGSB framework is not
appropriate in the adiabatic limit ($\beta \gg 1$).

@@

We mostly focused on vertically isothermal disks ($\Gamma=1$), since
this is the relevant case for protoplanetary disks that are irradiated
\citep{chiang97}. It also permits a largely analytical discussion.
Our analysis is an extension of the linear models considered in
\cite{lubow93} by  adding vertical shear and (rapid) thermal
relaxation, but in the limit of low-frequency perturbations in
nearly-Keplerian disks. These approximations were made for simplicity,
but also motivated by recent numerical simulations
\citep{nelson13}. We obtained explicit solutions for the VSI with
thermal relaxation in the thin disk limit.

@@

The fundamental VSI mode is characterized by a vertical velocity
perturbation with zero nodes and $\delta v_z\sim\mathrm{constant}$
near the mid-plane, and by one node in the density perturbation at the
midplane  (i.e. anti-symmetric about $z=0$).  These have been referred
to as `corrugation modes' by \cite{nelson13}, and were found to
dominate their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
protoplanetary disk model with thermal relaxation based on realistic
disk properties, we find the fundamental mode to be dominant and
robust against boundary conditions. The fundamental mode is thus
likely the most relevant VSI mode in protoplanetary disks.

@@

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we relax all the
approximations made in our analytical discussion, and accounted for
the background radial disk structure for consistency.  Our
numerical solutions are mostly consistent with our analytic 
expectations, including the dependence of $\beta_\mathrm{crit}$ on
disk parameters.

@@

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to
thermal relaxation. When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae
(MMSN). We consider the disk model described in \cite{chiang10}, and
modeled the thermal relaxation timescale $\beta$ based on
dust opacity, which results in a $\beta$ that depends on both the
perturbation lengthscale as well as the height from the midplane.  
We find the MMSN is unstable to the fundamental VSI at $10$s of AU,
with characteristic growth times of $\sim 30$ orbits. Although $\beta$
is not spatially constant as assumed in our analysis, we find that
evaluating $\beta < \beta_\mathrm{crit}$ at the disk
midplane gives a rough measure of the radii at which the VSI operates in the
MMSN.  This simple criteron may be a useful first step to 
estimate the relevance of the VSI in realistic astrophysical
disks without resorting to non-linear numerical simulations.

@@

\emph{Improved thermal evolution.} We assumed non-adiabatic effects
can be represented by a thermal relaxation timescale $t_c$. For 
application to protoplanetary disks we connected $t_c$ to the disk
properties, but one can improve the realism by generalizing the
opacity model adopted here to be density-dependent,  i.e. $\kappa_d =
\kappa_d(\rho, T)$.  However, going beyond the framework of thermal
relaxation requires a proper treatment for radiative diffusion, which
involves the Lapacian operator and additional boundary conditions
must be supplemented. We will consider these improvements in a
follow-up study.

@@

We have neglected viscosity. \cite{nelson13} have shown that an
alpha viscosity of $\lesssim O(10^{-4})$ is required for the VSI to
operate. It is of interest to quantify the effect of viscosity
on the linear VSI. This would help identify, more accurately, the
appropriate regions in a protoplanetary disk that are sufficiently
laminar to develop the VSI.  This problem is, however, complicated by
the fact that proper inclusion of viscosity introduces viscous heating
in the energy equation, as well as a meridional flow in the equilibrium state.

>>>>

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach to model thermal relaxation was also
employed in the global numerical simulations of \cite{nelson13}, and
allows the thermodynamic response of the disk to be controlled by
varying the parameter $\beta$. However, it is possible to model
$\beta$ self-consistently by relating it to thermal timescales
expected in realistic disks.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically-global shearing box' (VGSB) formalism developed by
\cite{mcnally14}, who showed the linear VSI studied by
\cite{nelson13} in a locally isothermal disk can be captured by the
VGSB. We find the VGSB framework is appropriate for non-isothermal
perturbations provided $\beta\lesssim1$.  
This is convenient because rapid thermal relaxation ($\beta\ll1$) is
generally required for the VSI.  However, the VGSB framework is not
appropriate in the adiabatic limit ($\beta \gg 1$).

@@

We mostly focused on vertically isothermal disks ($\Gamma=1$), since
this is the relevant case for protoplanetary disks that are passively irradiated by its central star 
\citep{chiang97}. It also permits a largely analytical discussion.
Our analysis is an extension of the linear models considered in
\cite{lubow93} by  adding vertical shear and (rapid) thermal
relaxation, but in the limit of low-frequency perturbations in
nearly-Keplerian disks. These approximations were made for simplicity,
but also motivated by recent numerical simulations
\citep{nelson13}. We obtained explicit solutions for the VSI with
thermal relaxation in the thin disk limit.

@@

The fundamental VSI mode is characterized by a vertical velocity
perturbation with zero nodes and $\delta v_z\sim\mathrm{constant}$
near the mid-plane, and by one node in the density perturbation at the
midplane  (i.e. anti-symmetric about $z=0$).  These have been referred
to as `corrugation modes' by \cite{nelson13}, and were found to
dominate their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
protoplanetary disk model with thermal relaxation based on realistic
disk properties, we find the fundamental mode to be the dominant mode,
 and it is robust against boundary conditions. The fundamental mode is
 thus likely the most relevant VSI mode in protoplanetary disks.

@@

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we relax all the
approximations made in our analysis, and accounted for
the background radial disk structure for consistency.  Our
numerical solutions are mostly consistent with our analytic 
expectations, including the dependence of $\beta_\mathrm{crit}$ on
disk parameters.  %for regimes of interest

@@

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI is insensitive to
thermal relaxation. When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

@@

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks, using a model for the the Minimum Mass Solar Nebulae
(MMSN) described in \cite{chiang10}. We  
modeled the thermal relaxation timescale $\beta$ based on realistic
disk properties and dust opacity, which results in a $\beta$ that
depends on both the perturbation lengthscale as well as the height
from the midplane. We find the MMSN is unstable to the VSI
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. Although $\beta$ is not vertically constant in the
MMSN as assumed in our analysis, we find that evaluating $\beta <
\beta_\mathrm{crit}$ at the disk midplane gives a rough measure of the
radii at which the VSI operates in the MMSN.  This simple criteron may
be a useful first step to estimate the relevance of the VSI in
realistic astrophysical disks without resorting to non-linear
numerical simulations.

@@

\emph{Improved thermal evolution.} We assumed non-adiabatic effects
can be represented by a thermal relaxation timescale $t_c$. For 
application to protoplanetary disks we connected $t_c$ to the disk
properties, but one can improve the realism further by generalizing the
opacity model adopted here to be density-dependent,  i.e. $\kappa_d =
\kappa_d(\rho, T)$ as described in \cite{bell94}. %two temperature
% structure  
However, going beyond the framework of thermal
relaxation requires a proper treatment for radiative diffusion, which
involves the Lapacian operator and additional boundary conditions
must be supplemented. We will consider these improvements in a
follow-up study.

@@

We have neglected viscosity. \cite{nelson13} have shown that an
alpha viscosity of $\lesssim O(10^{-4})$ is required for the VSI to
operate. It is of interest to quantify the effect of viscosity
on the linear VSI. This would help identify, more accurately, the
appropriate regions in a protoplanetary disk that are sufficiently
laminar to develop the VSI.  This problem is, however, complicated by
the fact that proper inclusion of viscosity introduces a meridional
flow in the equilibrium state, as well as viscous heating
in the energy equation.
